### 2018-08-13 response/response.tex

\documentclass[]{article}

\usepackage{booktabs} % For formal tables
\begin{document}

\title{Response}
\date{}
\author{}

\maketitle

%% We thank all the reviewers for their advices.

\section{Response to Reviewer 1}

\begin{verbatim}
Although the explanation of the problem and the solution is very
clear in general, I think that it would be useful to include an
explanation of how the mapping from virtual to physical addresses
could create additional conflicts in the shared cache, and an analysis
of whether this could happen in the target system or not.
\end{verbatim}

CPU caches can be either virtually indexed or physically indexed.
Nowadays, most processors use physically indexed caches,
that is, the physical address determines where the data should
be placed in the cache.
For physically indexed caches, SCP allocates physically continuous
memory buffers (using the \texttt{HUGETLBFS} filesystem provided
by the Linux kernel)
and partitions the buffer among threads sharing the same cache. 
For virtually indexed caches, virtually continuous
memory buffers (allocated with malloc)
are used instead of physically continuous ones.
Under both situations, in-cache matrix data of different threads
are guaranteed to be isolated from each other
and memory mapping from virtual to physical addresses
would not introduce additional conflicts.

Explanation on this problem has been added to the first paragraph
of section 4 (at the end of page 10).

\section{Response to Reviewer 2}

\begin{verbatim}
Why are the selected matrix sizes (3072 to 4096-6144) chosen?
\end{verbatim}

These sizes are large enough to make GEMM run stably at peak performance
given the total number of threads participating in the computation,
that is, 3072--4096 for 4 threads and 4096--6144 for 8, 16, 32 and 64 threads.

\begin{verbatim}
Which is the behavior of the approach for smaller matrices?
\end{verbatim}

The performance of GEMM on small matrices may show sub-optimal
and unstable performance.
It is common practice for research on GEMM optimization to omit small
matrices as GEMM is meant to run on large ones.
There is another variant of BLAS specially designed for
large amounts of small matrices,
Batched-BLAS, which is out of the scope of our paper.

\begin{verbatim}
Why is there a systematic drop for the largest tested matrices on
Figs. 5 and 6?
\end{verbatim}

There are two kinds of performance drop on Fig.~5 and Fig.~6.

(1) Performance drops as matrix size increases with $NT \ge 16$.
There is a synchronization point at the end of each iteration
of the loop at layer 2 (Fig.~2).
The synchronization is an all-to-all style
communication, whose overhead is proportional to $NT^2$.
As matrix size grows, GEMM suffers more synchronization because
the loop at layer 2 runs more iterations.
When $NT$ is 4 or 8, overhead of this synchronization is small,
but it increases quickly as $NT$ increases.

(2) Performance drops at matrix sizes which are multiple of 1024,
e.g., 3072, 4096, 5120, 6144. 
After profiling, we find that at these sizes,
GEMM suffers more memory contention.
As an example, the table below shows the profiling results
under matrix sizes 3968 and 4096
with parallelism configuration $NT_C=4$ and $NC=4$.
$BW_{A}$ and $BW_{B}$ are measured bandwidth of packing A and B, respectively.
$E_{avg}$ is the floating-point efficiency.
From the table we can see that there is a big performance gap
in $BW_{A}$ and $BW_{B}$.
$E_{avg}$ is also a little lower at 4096 than at 3840,
which may also be caused by the memory contention.
In this example, $NT=NT_C \times NC=16$.
When $M=N=K=4096$, each thread packs a $256 \times 256$ submatrix
from $A$ and a $256 \times 256$ submatrix from $B$,
both of size $V=512KB$.
When $M=N=K=3840$, the submatrix to be packed are of size $V=480KB$.
If the 16 threads starts packing roughly at the same time,
they will access memory addresses
$$addr+0V, addr+1V, \cdots, addr+15V$$
Due to the organization of DDR memory hardware,
$V=512K$ is more likely to cause memory contention than $V=480KB$
because the memory banks are accessed in an imbalanced manner.

\begin{table}[h]
  \centering
  \caption{Profiling results with $NT_C=4$ and $NC=4$}
  \vspace{1em}
  \begin{tabular}{cccc}
    \toprule
         & $BW_A$   & $BW_B$   & $E_{avg}$\\
    \midrule
    3840 & 0.81GB/s & 0.44GB/s & 87.2\%   \\
    4096 & 1.44GB/s & 0.89GB/s & 86.1\%   \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{verbatim}
Is there any evidence that this methodology transparently applies to 
other BLAS-3 routines in OpenBLAS (it should be trivial to test) and
for all matrix layouts (transposed/non transposed)?
\end{verbatim}

As other level-3 routines, for instance, TRSM, TRMM, SYMM and HEMM,
call GEMM internally, they should also benefit from our proposal.
So the SCP methodology can applies transparently to other BLAS-3 routines.

Internally, different matrix layouts (row-major/column-major)
are canonicalized to an implementation defined form
by optionally swapping the $A$ and $B$ operands in $C=\alpha A B+\beta C$.
For example, row-major $C=\alpha A B+\beta C$ can be transformed to
column-major $C^T=\alpha B^T A^T + \beta C^T$.
Then by data packing, matrix orientation information (transposed/non-transposed)
is eliminated out and matrix data is packed into the special
in-memory representation used by the kernel routine, as demonstrated in Fig. 2.
SCP works after these transformations and does not rely on
specific matrix layout or orientations.
So the SCP methodology applies transparently to
all matrix layouts and orientations.

\section{Response to Reviewer 3}

\begin{verbatim}
There is a typo on Page 7, line 43, turning -> tuning.
\end{verbatim}

Already fixed in the paper.

\begin{verbatim}
As the authors mentioned the related work on the tuning implementation
of GEMM only, they should also review the work regarding cache 
partitioning because this paper discussion is mainly on cache
partitioning. Some important papers are: ...
\end{verbatim}

The recommended citations are added to the ``related work`` section.

\begin{verbatim}
The proposed method works well, but it is so specific to DGEMM and
Phytium 2000+ processor. To confirm that their approach is much more
widely applicable even in the context of DGEMM, the discussion when
using the different types of processors with a cache hierarchy for
multi-core environments such as Xeon processors and KNL is strongly
required.
\end{verbatim}

While the Xeon and Xeon Phi processors series
have complex memory hierarchies,
all their caches use an LRU replacement policy.
So they are not target processors of our SCP method,
which must have shared, non-LRU caches.
%% Xeon processors have private, LRU L1 and L2 caches,
%% and a shared LRU last-level-cache (L3).
%% Xeon Phi (KNC, KNL) processors have LRU L1 and L2 caches.
As a result, evaluation on Xeon and Xeon Phi processors would not
improve the confirmability.

The SCP method is specific to GEMM, but its benefits can transparently
propagates to other BLAS-3 operations.
SCP is not specific to the Phytium 2000+ processor and
should work on any processors with shared non-LRU caches.
Unfortunately, we have not get another processor with shared
non-LRU caches like Phytium 2000+.

\end{document}

### 2018-08-13 response/response.tex

Thanks to all the reviewers for their comments.

\section{Response to Reviewer 1}

\begin{verbatim}
Although the explanation of the problem and the solution is very
clear in general, I think that it would be useful to include an
explanation of how the mapping from virtual to physical addresses
could create additional conflicts in the shared cache, and an analysis
of whether this could happen in the target system or not.
\end{verbatim}

Added. See the first block of sentences in 
\textcolor{blue}{blue} 
in Section 4.

@@

\section{Response to Reviewer 2}

\begin{verbatim}
Why are the selected matrix sizes (3072 to 4096-6144) chosen?
\end{verbatim}

Added. See the second block of sentences in 
\textcolor{blue}{blue} 
in Section 4.

@@

\begin{verbatim}
Which is the behavior of the approach for smaller matrices?
\end{verbatim}

The performance of GEMM on small matrices may be
sub-optimal
and unstable.
It is common to omit small
matrices as GEMM is meant to run on large ones.
There is a variant of BLAS specially designed for
large amounts of small matrices,
Batched-BLAS, which is out of the scope of this paper.

@@

\begin{verbatim}
Why is there a systematic drop for the largest tested matrices on
Figs. 5 and 6?
\end{verbatim}

There are two kinds of performance drops here:

(1) {\bf Performance drops as the matrix size increases 
under $NT \ge 16$.}
There is a synchronization point at the end of each iteration
of the loop at layer 2 (Fig.~2).
The synchronization is an all-to-all 
communication, whose overhead is proportional to $NT^2$.
As the matrix size increases, GEMM suffers
from more synchronization because
the loop at layer 2 runs for more iterations.
When $NT$ is 4 or 8, the overhead is small but
increases quickly as $NT$ increases.

(2) {\bf Performance drops at matrix sizes that are
multiples of 1024.}
After some  profiling, we found that at these sizes,
GEMM suffers from more memory contention.
As an example, the table below shows the profiling results
for matrix sizes 3968 and 4096
under $NT_C=4$ and $NC=4$.
$BW_{A}$ ($BW_{B}$) is the measured bandwidth for 
packing A (B), and
$E_{avg}$ is the floating-point efficiency.
We can see that there is a big performance gap
in $BW_{A}$ and $BW_{B}$.
$E_{avg}$ is also a little lower at 4096 than at 3840,
which may also be caused by the memory contention.
In this example, $NT=NT_C \times NC=16$.
When $M=N=K=4096$, each thread packs a $256 \times 256$ submatrix
from $A$ and a $256 \times 256$ submatrix from $B$,
both of size $V=512KB$.
When $M=N=K=3840$, the submatrix to be packed is 
of size $V=480KB$.
If the 16 threads start packing roughly at the same time,
they will access the following memory addresses:
$$addr+0V, addr+1V, \cdots, addr+15V$$
Due to the organization of the DDR memory hardware,
$V=512K$ is more likely to cause memory contention than $V=480KB$
because the memory banks are accessed in a non-balanced manner.

@@

\begin{table}[h]
  \centering
  \caption{Profiling results with $NT_C=4$ and $NC=4$}
  \vspace{1em}
  \begin{tabular}{cccc}
    \toprule
         & $BW_A$   & $BW_B$   & $E_{avg}$\\
    \midrule
    3840 & 0.81GB/s & 0.44GB/s & 87.2\%   \\
    4096 & 1.44GB/s & 0.89GB/s & 86.1\%   \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{verbatim}
Is there any evidence that this methodology transparently applies to 
other BLAS-3 routines in OpenBLAS (it should be trivial to test) and
for all matrix layouts (transposed/non transposed)?
\end{verbatim}

As other level-3 routines, e.g., TRSM, TRMM, SYMM and HEMM,
call GEMM internally, they should also benefit from our work.
So the SCP methodology can apply transparently to 
these other BLAS-3 routines as well.

Internally, different matrix layouts (row-major/column-major)
are canonicalized to an implementation defined form
by optionally swapping the $A$ and $B$ operands in $C=\alpha A B+\beta C$.
For example, the row-major $C=\alpha A B+\beta C$ can be transformed to
the column-major $C^T=\alpha B^T A^T + \beta C^T$.
Then by means of data packing, the
matrix orientation information (transposed vs. non-transposed)
is eliminated and the matrix data are
packed into the special
in-memory representation used by the kernel routine, as demonstrated in Fig. 2.

SCP, which works after these transformations, 
does not rely on
specific matrix layouts or orientations.

@@

\begin{verbatim}
There is a typo on Page 7, line 43, turning -> tuning.
\end{verbatim}

Fixed.

\begin{verbatim}
As the authors mentioned the related work on the tuning implementation
of GEMM only, they should also review the work regarding cache 
partitioning because this paper discussion is mainly on cache
partitioning. Some important papers are: ...
\end{verbatim}

The recommended references are now cited and
discussed in Section 5 (highlighted
in \textcolor{blue}{blue}).

@@

\begin{verbatim}
The proposed method works well, but it is so specific to DGEMM and
Phytium 2000+ processor. To confirm that their approach is much more
widely applicable even in the context of DGEMM, the discussion when
using the different types of processors with a cache hierarchy for
multi-core environments such as Xeon processors and KNL is strongly
required.
\end{verbatim}

For Xeon and Xeon Phi processors, the caches with
the LRU replacement policy are used.
However,
SCP is designed for processors with non-LRU caches.

SCP works for all other BLAS-3 routines, e.g., TRSM, TRMM, SYMM and HEMM, that call GEMM internally.
SCP is not specific to the Phytium 2000+ processor and
should work for any processors with shared non-LRU caches.

### 2018-08-13 evaluation.tex
Response revised by Prof Xue

We have implemented SCP in the OpenBLAS~\cite{openblas} library
%% (version 0.3.0-dev)
and evaluated it on a Phytium 2000+ processor.
The Phytium 2000+ processor is an emerging high-performance
64-core processor based on ARM's AArch64 architecture.
The 64 cores are organized into 16 clusters with each
cluster containing four cores.
The structure of a cluster is almost the same as
the four-core processor in Figure~\ref{fig:hierarchy}
except that the L2 caches of all 16 clusters connect to the main memory
with hardware coherence.
Table~\ref{tab:cluster} lists the
architectural parameters for the cores and caches
in this processor
and Table~\ref{tab:software} shows the software environment used.
We restrict our evaluation to DGEMM,
as in prior work~\cite{blispar,augem,poetmicro}, for two reasons.
First, the basic idea behind SCP applies to other
variants of GEMM such as SGEMM, CGEMM and ZGEMM.
Second, the LINPACK benchmark, which is used to build the
TOP500 list of world's most powerful supercomputers~\cite{top500},
relies on the DGEMM variant.

@@

>>>>

We have implemented SCP in the OpenBLAS~\cite{openblas} library
%% (version 0.3.0-dev)
and evaluated it on a Phytium 2000+ processor.
The Phytium 2000+ processor is an emerging high-performance
64-core processor based on ARM's AArch64 architecture.
The 64 cores are organized into 16 clusters with each
cluster containing four cores.
The structure of a cluster is almost the same as
the four-core processor in Figure~\ref{fig:hierarchy}
except that the L2 caches of all 16 clusters connect to the main memory
with hardware coherence.
Table~\ref{tab:cluster} lists the
architectural parameters for the cores and caches
in this processor
and Table~\ref{tab:software} shows the software environment used.
\textcolor{blue}{
The L2 caches in the Phytium 2000+ processor are physically indexed.
In our implementation, physically continuous
memory buffers (allocated by the \texttt{HUGETLBFS} filesystem
provided by the Linux kernel) are used for storing
matrices in packed form. Therefore,
for a given shared L2 cache,
the matrix data accessed by different threads 
are guaranteed to be free of conflict cache misses.
}
We restrict our evaluation to DGEMM,
as in prior work~\cite{blispar,augem,poetmicro}, for two reasons.
First, the basic idea behind SCP applies to other
variants of GEMM such as SGEMM, CGEMM and ZGEMM.
Second, the LINPACK benchmark, which is used to build the
TOP500 list of world's most powerful supercomputers~\cite{top500},
relies on the DGEMM variant.

@@

\textcolor{blue}{
The mairix sizes selected range from 3072 to 
4096 when $NT \in \{4, 8\}$ and to 
6144 when $NT \in \{16, 32, 64\}$. 
These sizes are large enough to make GEMM run stably at peak performance
the given the total number of threads participating in the computation.
}

### 2018-08-13 related.tex
Response revised by Prof Xue

>>>>

\textcolor{blue}{
SCP is a cache partitioning technique.
While SCP is specific to GEMM running on architectures
with non-LRU caches,
there are general cache partitioning techniques \cite{cp0, cp1, cp2}
that focus on estimating the interference across 
the multiple applications
sharing the same cache and improving the
overall system performance.
In these research efforts, either hardware modifications are required
or some sophisticated cache replacement policies are 
needed.
In contrast,
SCP is a pure software approach that works for 
commercial processors.
}
